---
title: "DATA 505: Lab 5"
subtitle: "Data manipulation & visualization"
date: "2025-09-23"
author: "Hien Bui"
format:
  html:
    toc: true
    toc-depth: 1
    number-sections: true
  pdf:
    toc: true
    toc-depth: 1
    number-sections: true
execute:
  eval: true
---

## Setup & Goals

Goals for Lab 5:

- Explore the [Changing Childhood Project](https://changingchildhood.unicef.org/) survey data
- Extract information, re-organize, and summarize data with tidyverse package functions
- Visualize data to generate informative, accurate, and high-quality visualizations

## The data

- The survey was done in two waves: phase one (2021) includes participants of 21 countries, and phase two (2022-2023) expands to 55 countries, but with a subset questions of phase one.  We'll be using the **phase two** data.
- Go to [this](https://www.unicef.org/innocenti/reports/tumultuous-world-through-childrens-eyes) page and click **Microdata (csv, 5 MB)** to download the data.  Double check that the resulting file is named *UNICEF-Changing-Childhood-Project-data-2022-2023*
- To make sense of the variables, we need these two code books that are created for the two survey waves.  If you cannot find a variable in one of them, try the other one. 
    + On the [same](https://www.unicef.org/innocenti/reports/tumultuous-world-through-childrens-eyes) page, download **Methodology and codebase (pdf, 1 KB)**
    + From [this](https://www.unicef.org/innocenti/reports/changing-childhood-project-report) page, download **Codebook - English (pdf, 376 KB)**
- Also, make sure you have loaded required packages
```{r}
# install the MASS package if needed
if(!"tidyverse" %in% installed.packages()){
  install.packages("tidyverse")
}

# load the MASS package and verify it is now ready for use
stopifnot(require(tidyverse))
```
```{r}
library(tidyverse)
```
- Put the data and the two pdf codebook files in the same folder as this qmd file.
- Import the data and save it as **df.child** with the following code.

```{r}
df.child <- read.csv("UNICEF-Changing-Childhood-Project-data-2022-2023.csv")
```


## Getting acquainted with the data

- Use the `str()` function to print a concise summary of the `df.child` object

```{r}
str(df.child)
```

- Call the `str()` function with the pipe operator

```{r}
df.child |> str()
```

- Use the `unique()` function to get a list of all the countries included in the dataset.

```{r}
df.child$countrynew |> unique()
```



## Basic univariate analysis

### WP1220
- Look up **WP1220** in the codebooks.  What does this variable represent?  
    - Please tell me your age. : It's the response's age
    
- Create a histogram of **WP1220** with base R.  Change at least one aesthetics.  Briefly discuss what the histogram shows.
   - Histogram shows all age frequency
   - The people <= 40yrs has a higher frequency than people > 40yrs
```{r}
hist(df.child$WP1220, 
     xlab = "Age", 
     main = "Histogram of WP1220",
     col = "lightblue")
``` 

- Create a histogram of **WP1220** with ggplot.  The aesthetics should look polished.
```{r}
ggplot(data = df.child,  aes(x = WP1220)) + 
  geom_histogram(binwidth=5,
                 fill="lightblue",
                 color="black") +
  theme_bw() +
  ggtitle('Histogram of WP1220') +
  labs(x='Age', y='Count')
``` 


### EMP_2010

- Look up **EMP_2010** in the codebooks.  What does this variable represent?  How many categories does it have?
    - EMP_2010 is Employment status. 
    - There are 6 categories:
      1. Employed full time for an employer 
      2. Employed full time for self 
      3. Employed part time do not want full time 
      4. Unemployed 
      5. Employed part time  
      6. want full time, Out of workforce

- Use the `table()` function to get a count of each category.

```{r}
table(df.child$EMP_2010)
```

- Convert the counts to percentages.

```{r}
round(prop.table(table(df.child$EMP_2010)),2)
```

- Create a bar graph of **EMP_2010** using ggplot.   The aesthetics should look polished.

```{r}
ggplot(data = df.child, 
       aes(x=EMP_2010)) +
  geom_bar(fill="lightblue",
                 color="black") + 
  theme_minimal() +
  ggtitle('Bar graph of EMP_2010') +
  labs(x='Employment', y='Count')
```

- Based on the information obtained above, briefly discuss your findings related to the variable **EMP_2010**.

   -  The highest count is want full time, Out of workforce (over 20000), the second count is Employed full time for an employer (over 10000), and the third count is Employed full time for self (almost 10000)
   
   
## Information extraction

### Compare age by country (complete example)

We want to compare participants' age by country using side-by-side boxplots.  This is difficult to show on one figure since there are many countries.  Instead, we subset to a few countries of interest.

```{r}
df.child |>
  filter(countrynew %in% c("Japan","United States","Yemen")) |>
  ggplot(aes(x=countrynew, y=WP1220)) +
  geom_boxplot() +
  labs(x='Country', y='Age') +
  theme_bw()
```


### Compare employment categories by country (complete example)

We want to compare employment categories (**EMP_2010**) by country.  Specifically, let's focus on 6 ("want full time, out of workforce").

```{r}
df.child |>
  group_by(countrynew) |>
  summarise(perc6 = mean(EMP_2010==6)) |>
  arrange(desc(perc6))
```


-  In the code above, the `mean` function is used to calculate the percentage.  Why can we do this?
    -  ***YOUR ANSWER***



## Open-ended data exploration and visualization

For this part, find variables you are interested in and create THREE or more interesting visualizations.  They should be varied (i.e., you should not create three histograms with three different variables) and should look polished.  Each visualization should involve a data manipulation step (such as subsetting or summarizing).  At least TWO of your visualizations should be done using ggplot.  For each visualization, write a brief description of what the graph shows.  

- I will color-print your visualizations and post it on the wall next to my office.
- For inspirations, you can take a look at [this](https://www.nytimes.com/2022/02/03/learning/whats-going-on-in-this-graph-feb-9-2022.html) or [this](https://www.nytimes.com/2021/11/17/upshot/global-survey-optimism.html) New York Times article based on the same data.
- There are many online resources for ggplot.  For example: search ggplot examples.
- Have fun and surprise me!


WP1219: Gender (1 Male 2 Female )
WP30: Standard of Living (1 Satisfied, 2 Dissatisfied, 3(DK), 4(Refused) )
WP3117 Education Level (1 Primary, 2 Secondary, 2 Tertiary, 4 (DK), 5 (RF) )

```{r}
names(df.child)
```

```{r}
df.child |>
  group_by(countrynew) |>
  summarise(satisf_rate = mean(WP30 == 1, na.rm = TRUE)) |>
  arrange(desc(satisf_rate)) |>
  head(n = 20) |>
  mutate(highlight = ifelse(row_number() == 1, "Top", "Other")) |>
  ggplot(aes(x = reorder(countrynew, satisf_rate), y = satisf_rate, fill = highlight)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Top" = "lightblue", "Other" = "gray")) +
  labs(x = "Country", y = "Proportion Satisfied (WP30 == 1)",
       title = "Top 20 Countries by Life Satisfaction") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_y_continuous(
    limits = c(0, 1),                     # force exactly 0–1
    expand = c(0, 0)                        # remove extra white space
  )
```
  
  - Overall: All top-20 countries are high—everyone is roughly 0.70–0.95 satisfied.
  - Leaders: Vietnam is clearly the highest; Tajikistan follows; Germany and the UK are also very strong.
  - Middle cluster: Many countries (e.g., Malaysia, France, Spain, Mexico, Indonesia, Philippines) sit tightly in the ~0.75–0.85 band.

```{r}
perc_table <- df.child |>
  filter(!is.na(WP30)) |>
  group_by(countrynew) |>
  summarise(
    satisfied    = mean(WP30 == 1),
    dissatisfied = mean(WP30 == 2),
    .groups = "drop"
  ) |>
  mutate(
    # now round for reporting
    satisfied    = round(satisfied * 100, 2),
    dissatisfied = round(dissatisfied * 100, 2),
    dk_refused   = round(100 - satisfied - dissatisfied, 2) 
  ) |>
  arrange(desc(satisfied)) |>
  head(20)

perc_table

```

```{r}
perc_table_long <- perc_table |>
  select(countrynew, satisfied, dissatisfied, dk_refused) |>
  pivot_longer(
    cols = c(satisfied, dissatisfied, dk_refused),
    names_to = "response",
    values_to = "percent"
  )
perc_table_long

# bring 'satisfied' into the long table
plot_df <- perc_table_long |>
  left_join(select(perc_table, countrynew, satisfied), by = "countrynew")

# (optional) keep top 20 by satisfied
top20 <- plot_df %>%
  distinct(countrynew, satisfied) |>
  arrange(desc(satisfied)) |>
  head(20) |>
  pull(countrynew)

plot_df <- filter(plot_df, countrynew %in% top20)
```
```{r}
ggplot(plot_df, aes(x = reorder(countrynew, satisfied), y = percent, fill = response)) +
  geom_col(position = "stack") +
  coord_flip() +
  labs(
    x = "Country",
    y = "Percentage",
    fill = "Response",
    title = "Top 20 Countries by Satisfaction Levels"
  ) +
  scale_fill_manual(values = c(
    "satisfied" = "lightblue",
    "dissatisfied" = "steelblue",
    "dk_refused" = "darkgrey"
  )) +
  scale_y_continuous(
    limits = c(0, 100),                     # force exactly 0–100
    expand = c(0, 0)                        # remove extra white space
  ) +
  theme_bw()
```

  
  - Overall: All top-20 countries are high—most have ≥75% satisfied.
  - Leaders: Vietnam is highest; Tajikistan, Germany, UK close behind.  
  - Lagging within top-20: Brazil, Nepal, South Korea show the largest dissatisfied slices.
  - DK/Refused: very small almost everywhere - differences come from satisfied vs. dissatisfied.
  - Spread: roughly mid-70s to mid-90s% satisfied across the group.
  
  

```{r}
world_map <- map_data("world")
world_map |> head(10)
```

```{r}
# compute share satisfied by country (fraction 0–1; use *100 if you want percent)
perc_table <- df.child |>
  group_by(countrynew) |>
  summarise(
    satisfied = mean(WP30 == 1, na.rm = TRUE),
    .groups = "drop"
  )
perc_table |> head(10)
```

```{r}
# join country polygons with your metric
map_data_55 <- world_map |>
  left_join(perc_table, by = c("region" = "countrynew"))
map_data_55 |> head(20)
```

```{r}
# check if country names match
setdiff(perc_table$countrynew, unique(world_map$region))
perc_table <- perc_table |>
  mutate(countrynew = case_when(
    countrynew == "United States" ~ "USA",
    countrynew == "United Kingdom" ~ "UK",
    TRUE ~ countrynew
  ))
```
```{r}
# join country polygons with your metric and check again
map_data_55 <- world_map |>
  left_join(perc_table, by = c("region" = "countrynew"))
setdiff(perc_table$countrynew, unique(world_map$region))
```

```{r}
head(map_data_55) #quick sanity check
```

```{r}
library(plotly)  # enables ggplot → interactive plot conversion via ggplotly()

# 1) Build tooltip text for hover (used only by Plotly)
map_data_55 <- map_data_55 |>
  mutate(
    # 'hover' becomes the value of the ggplot aesthetic 'text' below.
    # <br> adds a line break inside the tooltip.
    hover = paste0(region, "<br>Satisfied: ", round(satisfied, 2), "%")
  )

# 2) Static ggplot map (we convert this to Plotly afterward)
p <- ggplot(
  map_data_55,
  aes(
    long, lat,                  # polygon coordinates from map_data('world')
    group = group,              # keeps multi-part polygons (islands) together
    fill  = satisfied,          # metric mapped to the color scale
    text  = hover               # custom tooltip text (passed to Plotly)
  )
) +
  geom_polygon(color = "white") +  # draw country shapes with white borders
  scale_fill_gradient(
    low      = "lightcyan",        # color for low values
    high     = "steelblue4",       # color for high values
    na.value = "lightgrey",        # color for countries with no data
  ) +
  coord_fixed(1.4) +               # fix aspect ratio so lat/long aren’t stretched
  theme_void() +                   # remove axes, ticks, background (clean map style)
  labs(
    fill  = "Satisfied %"                # legend title for the fill scale
      ) 

ggplotly(p, tooltip = "text") |>
  layout(
    title = list(
      text = "Where Satisfaction Lives<br><sup>Gray = missing data; Source: UNICEF 2022–2023</sup>",
      x = 0.5 # center
    ),
    margin = list(t = 90) # t is top
  )

```


  - There are only 55 countries in this visualization.
  - Europe & North America: mostly higher satisfaction.
  - South America: several higher, but mixed across countries.
  - Asia: very mixed—East/Southeast tend mid-to-high; South/Central lighter.
  - Middle East & Africa: more light shades overall and many gray gaps.
  - Oceania: limited coverage; where shown, generally higher.
    
